--- 
title: "Statistical Modeling: A Fresh Approach (3/ed)"
author: "Daniel Kaplan"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::html_book:
    toc: no
    css: toc.css
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This is a minimal example of using the bookdown package to write a book. The output format for this example is bookdown::gitbook."
---

# To build this book

In the shell, give
```
sh _build.sh
```

To get the navigation as a bar, somehow contrive to put these two
lines just before the table of contents in the HTML file:
```
<link rel="stylesheet" href="style.css" type="text/css" />
<link rel="stylesheet" href="toc.css" type="text/css" />
```

For a PDF file, change `_build.sh`
```
Rscript -e "bookdown::render_book('index.Rmd', bookdown::tufte_book2())"

```
# Prerequisites

Can Danny 

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

For now, you have to install the development versions of **bookdown** from Github:

```{r eval=FALSE}
devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need to install XeLaTeX.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
