%% Author: Daniel Kaplan
%% Subject: Data organization (identifying cases and variables)

Here is a small data frame about automobiles.

\bigskip
\centerline{\small\begin{tabular}{|rrrrll|}\hline
Make and & Vehicle & Trans. & \# of & City & Hwy \\
model & type & type & cyl. & MPG & MPG \\ \hline \hline
Kia Optima & compact & Man. & 4 & 21 & 31\\
Kia Optima & compact & Auto. & 6 & 20 & 28\\
Saab 9-7X AWD & SUV & Auto. & 6 & 14 & 20\\
Saab 9-7X AWD & SUV & Auto. & 8 & 12 & 16\\
Ford Focus & compact & Man. & 4 & 24 & 35\\
Ford Focus & compact  & Auto. & 4 & 24 & 33\\
Ford F150 2WD & pickup  & Auto. & 8 & 13 & 17\\\hline
\end{tabular}}

\bigskip

\begin{enumerate}[(a)]
\item What are the cases in the data frame?  

\begin{MultipleChoice}
\wrong{Individual car companies}
\wrong{Individual makes and models of cars}
\correct{Individual configurations of cars}
\wrong{Different sizes of cars}
\end{MultipleChoice}

\begin{AnswerText}
Each case is a different make and model in a particular
configuration.  For instance, there are two cases that are Kia
Optimas.  One of these cases has a manual transmission, the other an
automatic transmission.  For the Saab, the two cases have different
numbers of cylinders, even though both are automatic transmissions.
\end{AnswerText}

\item For each case, what variables are given?  Are they categorical 
or quantitative?

\begin{itemize}
\item Kia Optima:
  \SelectSetHoriz{not.a.variable}{not.a.variable,categorical,quantitative}

\begin{AnswerText}
The Kia Optima is not a variable, it's a level of the \VN{Make and
  model} variable.
\end{AnswerText}

\item City MPG: \SelectSetHoriz{quantitative}{not.a.variable,categorical,quantitative}

\item Vehicle type: \SelectSetHoriz{categorical}{not.a.variable,categorical,quantitative}

\item SUV:  \SelectSetHoriz{not.a.variable}{not.a.variable,categorical,quantitative}

\begin{AnswerText}
SUV is a level of the \VN{Vehicle type} variable.
\end{AnswerText}

\item Trans. type: \SelectSetHoriz{categorical}{not.a.variable,categorical,quantitative}

\item \# of cyl.: \SelectSetHoriz{quantitative}{not.a.variable,categorical,quantitative}
\end{itemize}

\item Why are some cars listed twice?  Is this a mistake in the table?

\begin{MultipleChoice}
\wrong{Yes, it's a mistake.}
\correct{A car brand might be listed more than once, but the cases have
  different attributes on other variables.}
\wrong{Some cars are more in demand than others.}
\end{MultipleChoice}

\begin{AnswerText}
As explained above, the cases with the same make and model are
different in other variables.
\end{AnswerText}

\end{enumerate}

