%% Author: Daniel Kaplan
%% Subject: Models (introduction)

Many fields of natural and social science have principles that are
identified by name. Sometimes these are called ``laws,'' sometimes
``principles'', ``theories,'' etc.  Some examples:

\bigskip
\noindent\begin{tabular}{lll}
Kepler's Law & \multicolumn{2}{l}{Newton's Laws of Motion}\\
Ohm's Law & Grimm's Law & Nernst equation\\
Raoult's Law & Nash equilibrium & Boyle's Law\\
Zipf's Law & \multicolumn{2}{l}{Law of diminishing marginal utility} \\
Pareto principle & Snell's Law & Hooke's Law\\
Fitt's Law & \multicolumn{2}{l}{Laws of supply and demand}\\
Ideal gas law & \multicolumn{2}{l}{Newton's law of cooling}\\
\multicolumn{2}{l}{Le Chatelier's principle} & Poiseuille's law\\
\end{tabular}
\bigskip

These laws and principles can be thought of as models.  Each is a
description of a relationship.  For instance, Hooke's law relates the
extension and stiffness of a spring to the force exerted by the
spring.  The laws of supply and demand relate the quantity of a good
to the price and postulates that the market price is established at
the equilibrium of supply and demand.

Pick a law or principle from an area of interest to you --- chemistry,
linguistics, sociology, physics, ... whatever.  Describe the law, what
quantities or qualities it relates to one another, and the ways in
which the law is a model, that is, a representation that is suitable
for some purposes or situations and not others.

Enter your answer here: \TextEntry  

An example is given below.


\begin{quotation}
\em
EXAMPLE: As described in the text, Hooke's Law, $f = -k x$, relates the
force ($f$), the stiffness ($k$) and the extension past resting length ($x$)
for a spring.  It is a useful and accurate approximation for small
extensions.  For large extensions, however, springs are permanently
distorted or break.  Springs involve friction, which is not included
in the law.  Some springs, such as passive muscle, are really
composites and show a different pattern, e.g., $f = -  k \ {x^3} / {|x|}$ for
moderate sized extensions.
\end{quotation}


