There are many different numerical descriptions of distributions:
mean, median, standard deviation, variance, IQR, coverage interval,
...   And these are just the ones we have touched on so far.  We'll
also encounter ``standard error,''  ``margin of error,''  ``confidence
interval.''  There are so
many that it becomes a significant challenge to students to keep them
straight.  Eventually, statistical workers learn the subtleties of the
different descriptions and when each is appropriate.  Then, like using
near synonyms in English, it becomes second nature.  

As an example,
consider the verb ``spread.''.  Here are some
synonyms from the thesaurus, each of which is appropriate in a particular context:
broadcast, scatter, propagate, sprawl, extend, stretch, cover, daub, ...  If you
were talking to a farmer about sewing seeds, the words ``broadcast''
or ``scatter'' would be appropriate, but it would be silly to say the
seeds are being ``daubbed'' or ``sprawled''.  On the other hand, to an urbanite
concerned with congestion in traffic, the growth of the city might
well be summarized with ``sprawl.''  You have to know the context and
the intent to choose the correct term.  

To help to understand the different context and intents, here are
two important ways of categorizing what a particular description captures:

\begin{itemize}
\item[(1)] Location and scatter
\begin{itemize}
\item What is a typical value? (``center'')
\item What are the top and bottom range of the values? (``range'')
\item How far are the values scattered? (``scatter'')
\item What is high? or What is low? (``non-central'')
\end{itemize}

\item[(2)] Including the ``extremes''
\begin{itemize}
\item All inclusive, and sensitive to outliers. (``not-robust'')
\item All inclusive, but not sensitive to outliers. (``robust'')
\item Leaves out the very extremes. (``plausible''')
\item Focuses on the middle. (``mainstream'')
\end{itemize}
Note that descriptors of both the ``plausible'' and the ``mainstream''
type are necessarily robust, since they leave out the outliers.

\item[(3)] Individual versus whole sample.
\begin{itemize}
\item Description relevant to individual cases
\item Description or summary of entire samples, combining many cases.
\end{itemize}
You won't have to deal with this until later, where it explains terms
that you haven't yet encountered like like ``standard error'',
``margin of error'', ``confidence interval.''

\end{itemize}

\medskip

\noindent Example: The {\bf mean} describes the {\em center} of a
distribution.  It is calculated from all the data and {\em not-robust}
against outliers.

\bigskip

For each of the following descriptors of a distribution , choose the
items that best characterize the descriptor.

\begin{enumerate}
\item Median
\begin{enumerate}
\item \SelectSetHoriz{center}{center,range,scatter,non-central}
\item \SelectSetHoriz{robust}{robust,not-robust,plausible,mainstream}
\end{enumerate}

\item Standard Deviation
\begin{enumerate}
\item \SelectSetHoriz{scatter}{center,range,scatter,non-central}
\item \SelectSetHoriz{not-robust}{robust,not-robust,plausible,mainstream}
\end{enumerate}

\item IQR
\begin{enumerate}
\item \SelectSetHoriz{scatter}{center,range,scatter,non-central}
\item \SelectSetHoriz{robust}{robust,not-robust,plausible,mainstream}
\end{enumerate}

\item Variance
\begin{enumerate}
\item \SelectSetHoriz{scatter}{center,range,scatter,non-central}
\item \SelectSetHoriz{not-robust}{robust,not-robust,plausible,mainstream}
\end{enumerate}

\item 95\% coverage interval
\begin{enumerate}
\item \SelectSetHoriz{range}{center,range,scatter,non-central}
\item \SelectSetHoriz{plausible}{robust,not-robust,plausible,mainstream}
\end{enumerate}

\item 50\% coverage interval
\begin{enumerate}
\item \SelectSetHoriz{range}{center,range,scatter,non-central}
\item \SelectSetHoriz{mainstream}{robust,not-robust,plausible,mainstream}
\end{enumerate}

\item 50th percentile
\begin{enumerate}
\item \SelectSetHoriz{non-central}{center,range,scatter,non-central}
\end{enumerate}

\item 80th percentile
\begin{enumerate}
\item \SelectSetHoriz{non-central}{center,range,scatter,non-central}
\end{enumerate}

\item 99th percentile
\begin{enumerate}
\item \SelectSetHoriz{non-central}{center,range,scatter,non-central}
\end{enumerate}

\item 10th percentile
\begin{enumerate}
\item \SelectSetHoriz{non-central}{center,range,scatter,non-central}
\end{enumerate}

\end{enumerate}

One of the reasons why there are so many descriptive terms is that
they have different roles in theory.  For example, the variance turns
out to have simple theoretical properties that make it useful when
describing sums of variables.  It's much simpler than, say, the IQR.

