# Computational Technique

```{r include = FALSE}
library(mosaic)
detex_index_entry <- function(...) {NULL}
```


Calculating means and other simple statistics is a matter of using the
right function in R.  The **`mosaic`** package --- which you should
load in to R as shown in Section \@ref("sec:loading-mosaic") --- makes it straightforward to calculate either a "grand" statistic or a
"group-wise" statistic.   To illustrate:

Load the **`mosaic`** package, needed only once per session.
```{r, include = FALSE}
require(mosaic) 
```

Read in data you are interested in analyzing, for instance
the Cherry-Blossom 2008 data described earlier:^[These are also made available in the **`mosaic`** package as `TenMileRace`.]  

```{r}
Runners = read.csv("http://tiny.cc/mosaic/Cherry-Blossom-2008.csv")
names( Runners )
```

Calculate a grand mean on the "gun" time --- the time between
the start of the race, signalled by a gun, and when each runner
crossed the finish line.  Here, to emphasize that this is a group--wise model, we use the
function `gwm()` from the **`mosaic`** package.  It works the same as `mean()`, but its output
is more formally the output of a model.

```{r}
gwm(gun ~ 1, data = Runners)
```
The model formula `gun ~ 1` is an "all cases the same" model.  You will refer to the `1` as the "Intercept  term" later in the text, but for now, think of it as "one value."  The one value here is the grand mean.
Grand mean and other "grand" statistics.

```{r}
mean( gun ~ 1, data = Runners)
median( gun ~ 1, data = Runners )
sd( gun ~ 1, data = Runners )
```

To tell R that you want to break the statistics down by groups, replace the `1` with the variable
which defines the groups.  You will be
using this notation frequently in building models.  Here, as before, the `~` means, "model
by" or "broken down by" or "versus."  To find the means for the gun time broken down by sex, enter
```{r}
gwm( gun ~ sex, data = Runners )
```

Other statistics work the same way, for instance,
```{r}
sd( gun ~ sex, data = Runners )
```

Another example ... wage broken down by sector of the economy, using
data `cps.csv`:^[This data set is made available by **`mosaic`** as `CPS85`.]

```{r}
CPS = read.csv("http://tiny.cc/mosaic/cps.csv")
gwm( wage ~ sector, data = CPS )
```

In the Whickham smoking data example^[This data set is made available by **`mosaic`** as `Whickham`.], the `outcome` for each person was
not a number but a category: Alive or Dead at the time of the
follow-up.  

```{r}
W <- read.csv("http://tiny.cc/mosaic/whickham.csv")
names(W)
levels(W$outcome)
```

The `gwm()` function recognizes that `outcome` is a categorical variable and outputs proportions.  In the case of "all cases the same," this is porportions of the data in the `outcome` levels:

```{r}
gwm( outcome ~ 1, data = W )
```

Here's the breakdown of the levels `Alive` and `Dead` according to smoking status:
```{r}
gwm( outcome ~ smoker, data = W )
```

You may be surbrised by this result.  But don't be misled.
A more meaningful question is whether smokers are different from
non-smokers when holding other variables constant, such as age.  To
address this question, you need to add age into the model.

It might be natural to consider
each age --- 35, 36, 37, and so on --- as a separate group, but you
won't get very many members of each group.  And, likely, the data for
35 year-olds has quite a lot to say about 36 year-olds, so it doesn't
make sense to treat them as completely separate groups.

You can use the `cut()` function to divide up a quantitative
variable into groups.  You get to specify the breaks between groups.

```{r}
W$ageGroups <- cut(W$age, breaks=c(0,30,40,53,64,75,100) )
gwm( outcome ~ ageGroups, data = W )
gwm( outcome ~ smoker + ageGroups, data = W )
```

With modeling techniques, to be introduced in later chapters, you can use
quantitative variables without the need to divide them into groups.

## Model Values and Residuals

A group-wise model tells you a model value for each group, but often
you will need these in a case-by-case format: the model value for each
case in a data set.  The `fitted()` function carries out
this simple calculation, taking each case in turn, figuring out which
group it belongs to, and then returning the set of model values for
all the cases.  It requires two arguments: the model (here provided by `gwm()`)
the data on which the model was based.  For example:

`r detex_index_entry('P', 'fitted')` 
`r detex_index_entry('P', 'resid')` 

```{r}
Kids <- KidsFeet   # just get this from mosaicData instead of going out on the web with read.csv
mod <- gwm( width ~ sex, data = Kids )
fitted( mod, Kids )
```

The residuals are found by subtracting the case-by-case model value
from the actual values for each case.  

```{r}
res <- Kids$width - fitted(mod, Kids)
head(res) # display the first few residuals
res <- resid(mod, Kids) # function resid does the calculation automatically
head(res)
```

Take care to use the same quantitative variable (`width` in this case) from the
data as was used in constructing the model.

The `var()` function will calculate the variance:

```{r}
var( Kids$width )  # overall variation
var( fitted(mod, Kids) ) # variation in model values
var( resid(mod, Kids) ) # residual variation
```

Notice the "model triangle" relationship between these three numbers.
