# 9.6 Computational Technique

```{r, message=FALSE}
require(mosaic)
```

The coefficient of determination, $R^2$, 
compares the variation in the response variable to the variation
in the fitted model value.  It can be calculated as a ratio of variances: 

```{r}
Swim <- SwimRecords # from mosaicData
mod <- lm( time ~ year + sex, data = Swim)
var(fitted(mod)) / var(Swim$time)
```
`r detex::index_entry('C', 'coef. of determination!computing')` 
`r detex::index_entry('P', 'var@var')` 
`r detex::index_entry('P', 'Descriptive Statistics!var@var')` 

The convenience function `rsquared()` does the calculation for you:
```{r}
rsquared(mod)
```

The `r detex::new_word('regression report')` is a standard way of summarizing
models.  Such a report is produced by most statistical software
packages and used in many fields.  The first part of the table
contains the coefficients --- labeled "Estimate" --- along with
other information that will be introduced starting in Chapter
\@ref("chap:confidence").  The $R^2$ statistic is a standard part of the
report; look at the second line from the bottom.
```{r}
summary(mod)
```


Occasionally, you may be interested in the correlation coefficient $r$ between two
quantities.  
You can, of course, compute $r$ by fitting a model, finding $R^2$,
and taking a square root.
```{r}
mod2 <- lm( time ~ year, data = Swim)
coef(mod2)
sqrt(rsquared(mod2))
```

The `cor()` function computes this directly:
```{r}
cor(Swim$time, Swim$year)
```

Note that the negative sign on $r$ indicates that record swim `time`
decreases as `year` increases.  This information about the
direction of change is contained in the sign of the coefficient from
the model.  The magnitude of the coefficient tells how fast the
`time` is changing (with units of seconds per year).  The
correlation coefficient (like $R^2$) is without units.

Keep in mind that the correlation coefficient $r$ summarizes only the
simple linear model A ~ B where B is quantitative.  But the
coefficient of determination, $R^2$, summarizes any model; it is much
more useful.  If you want to see the direction of change, look at the
sign of the correlation coefficient.

