# 15.8 Computational Technique

```{r, message = FALSE}
require(mosaic) # mosaic operators and data used in this section
```


The basic software for hypothesis testing on parts of models involves the 
familiar `lm` and `summary` operators for generating the
regression report and the `anova` operator for generating an
**`ANOVA`**   report on a model.

## ANOVA reports

`r detex::index_entry('P', 'anova@anova')` 
`r detex::index_entry('P', 'Hypothesis Testing!anova@anova')` 

The `anova` operator takes a model as an argument and produces
the term-by term `ANOVA` report.  To illustrate, consider this model
of wages from the Current Population Survey data.  

```{r}
Cps <-  CPS85 # from mosaicData
mod1 <-  lm( wage ~ married + age + educ, data = Cps)
anova(mod1)
```

Note the small p-value on the `married` term: `0.0097`.

To change the order of the terms in the report, you can create a new
model with the explanatory terms listed in a different order.  For
example, here's the `ANOVA` on the same model, but with `married`
last instead of first:
```{r}
mod2 <- lm( wage ~ age + educ + married, data = Cps)
anova(mod2)
```

Now the p-value on `married` is large.  This suggests that much of
the variation in `wage` that is associated with `married` can
also be accounted for by `age` and `educ` instead.


## Non-Parametric Statistics

Consider the model of world-record swimming times plotted on page 116.

<!-- TEX COMMAND NOT FOUND pageref page:interaction-swimming .  -->

It shows pretty clearly the
interaction between `year` and `sex`.

It's easy to confirm that this interaction term is statistically significant:

```{r}
Swim <-  SwimRecords  # in mosaicData
anova( lm( time ~ year * sex, data = Swim) )
```

The p-value on the interaction term is very small: $2.8 \times10^{-6}$.

To check whether this result might be influenced by the shape of the
distribution of the `time` or `year` data, you can conduct a
non-parametric test.  Simply take the rank of each quantitative variable:


`r detex::index_entry('C', 'non-parametric statistics')` 
`r detex::index_entry('C', 'outlier!and non-parametrics')` 

`r detex::index_entry('P', 'rank@rank')` 
`r detex::index_entry('P', 'Hypothesis Testing!rank@rank')` 

```{r}
mod <- lm( rank(time) ~ rank(year) * sex, data = Swim)
anova(mod)
```

With the rank-transformed data, the p-value on the interaction term is
much larger: no evidence for an interaction between
`year` and `sex`.  You can see this directly in  a plot of the data
after rank-transforming `time`:

```{r}
xyplot( rank(time) ~ year, groups = sex, data = Swim)
```

The rank-transformed data suggest that women's records are improving
in about the same way as men's. 
That is, new records are set by women at a rate similar to the rate
at which men set them.
