# 12.7 Computational Technique

```{r, message=FALSE}
require(mosaic) # mosaic defined operators and data used below
```

Regression reports are generated using software you have already encountered: `lm` to fit a model and `summary` to construct the report from
the fitted model.  To illustrate:   
```{r}
Swim <- SwimRecords #from mosaicData
mod <- lm(time ~ year + sex, data = Swim)
summary(mod)
```


## Confidence Intervals from Standard Errors

`r detex::index_entry('P', 'Confidence Intervals')` 

Given the coefficient estimate and the standard error
from the regression report, the confidence interval is easily generated.  
For a 95\%  confidence interval, you just multiply the standard error by 2
to get the margin of error.  For example, in the above, the
margin of error on `sex M` is $2 \times 1.013 = 2.03$, or, in 
computer notation:
```{r}
2 * 1.0129
```

If you want the two endpoints of the confidence interval, rather than 
just the margin of error, do these
simple calculations: (1) subtract the margin of error from the estimate; 
(2) add the margin of error to the estimate.  So, 
```{r}
-9.798 - 2*1.0129
-9.798 + 2*1.0129
```

The key thing is to remember the multiplier that is applied 
to the standard error.  A multiplier of approximately 2 is for
a 95\% confidence interval.

The `confint()` function provides a convenient way to calculate
confidence intervals directly.

`r detex::index_entry('C', 'confidence interval!computation')` 
`r detex::index_entry('P', 'confint@confint')` 
`r detex::index_entry('P', 'Confidence Intervals!confint@confint')` 
It calculates the exact multiplier (which depends somewhat on the
sample size)  and applies it to the standard error to produce the confidence intervals.

```{r}
mod <- lm( time ~ year + sex, data = Swim)
confint(mod)
```

It would be convenient if the regression report included confidence
intervals rather than the standard error.  Part of the reason it
doesn't is historical: the desire to connect to the traditional by-hand calculations.



## Bootstrapping Confidence Intervals

Confidence intervals on model coefficients can be computed using the
same bootstrapping technique introduced in Chapter \@ref("chap:statistical-inference").

Start with your fitted model. To illustrate, here is a model of
swimming time over the years, taking into account sex:
```{r}


swim <- SwimRecords
lm( time ~ year + sex, data = swim)
```
These coefficients reflect one hypothetical draw from the
population-based sampling
distribution.  It's impossible to get another draw from the
``population" here: the actual records are all you've got.

But to approximate sampling variation, you can treat the sample as
your population and re-sample:
```{r}
lm( time ~ year + sex, data = resample(Swim))
```

Constructing many such re-sampling trials and collect the results 

```{r}
s = do(500) * lm( time ~ year + sex, data = resample(Swim))
head(s)
```

To find the standard error of the coefficients, just take the standard
deviation across the re-sampling trials.  For the indicator variable `sex M`:
```{r}
sd(s$sexM)
```

Multiplying the standard error by 2 gives the approximate 95\% margin
of error.  Alternatively, you can use the `confint()` function
to calculate this for you:
```{r}
confint(s$sexM, method = "stderr")
```



## Prediction Confidence Intervals

`r detex::index_entry('C', 'confidence interval!for prediction')` 
`r detex::index_entry('C', 'prediction!confidence interval')` 

When a model is used to make a prediction, it's helpful to be able to
describe how precise the prediction is.  For instance, suppose you want to use the `KidsFeet` data set to make a prediction of the foot width of a girl whose foot length is 25 cm. 

`r detex::index_entry('P', 'predict@predict')` 
`r detex::index_entry('P', 'Modeling!predict@predict')` 

First, fit your model:
```{r}
Kids = KidsFeet # from mosaicData
names(Kids)

levels(Kids$sex)

mod <- lm( width ~ length + sex, data = Kids)
```

Now apply the model to the new data for which you want to make a
prediction.  Take care to use the right coding for categorical variables.
```{r}
predict(mod, newdata = data.frame( length=25, sex="G" ))
```

In order to generate a confidence interval, the `predict`
operator needs to be told what type of interval is wanted.  There are
two types of prediction confidence intervals:
<!-- begin description -->

#. Interval on the model value which reflects the sampling
  distributions of the coefficients themselves.  To calculate this,
  use the `interval = "confidence"` named argument:
```{r}
predict(mod, newdata = data.frame( length = 25, sex = "G" ), 
         interval = "confidence")
```

The components named `lwr` and `upr` are the lower and upper
limits of the confidence interval, respectively.

`r detex::index_entry('P', 'interval@interval=!in predict')` 
`r detex::index_entry('P', 'Confidence Intervals!interval@interval=!in predict')` 

#. Interval on the prediction which includes the variation due to
  the uncertainty in the coefficients as well as the size of a typical
  residual.  To find this interval, use the
  `interval = "prediction"` named argument:
```{r}
predict(mod, newdata = data.frame( length = 25, sex = "G" ), 
         interval = "prediction")
```

The prediction interval is larger than the model-value confidence
interval because the residual always gives additional uncertainty
around the model value. Predicting an individual's foot width, even if we
know her sex and foot length, involves a lot more
uncertainty than predicting the mean foot width of all individuals with these
particular characteristics.
<!-- end description -->  


